# Final reports

## Overview

In Week 10 you will be submitting your final reports.

## Citing R and packages

The underlying programme is clearly a massive piece of work and you should give its authors credit by citing them in any work you do using R.

For example you might write:

> Statistical analysis was done using R 4.1.1 [@R-base].

R makes it easy to generate the right reference to use because there's a built-in function to do it.

```{r}
#| label: w10_cite
citation()
```

You should also cite the main packages you've used. For example:

> Data wrangling was carried out with `dplyr` [@R-dplyr] and other packages from the `tidyverse` [@tidyverse]; graphs were plotted using `ggplot2` [@R-ggplot2].

Again, you can use `citation` to generate the correct references. For example:

```{r}
#| label: w10_cite2
citation('dplyr')
```

If you are wondering, `BibTeX` is a file format (and software) used to describe lists of references, often for use within LaTeX documents. The `rmarkdown` package [@rmarkdown2018] provides methods to work with BibTeX references.

### What to cite?

You should always cite R itself, but there is an element of judgement in deciding which individual packages to cite. For this module, you won't lose marks as long as you have made a reasonable effort.

To make your work fully reproducible it is best practice to list all the packages you've used (e.g. in an appendix). You can generate the necessary information with the function `sessionInfo()`, but this is not necessary for FDS coursework.


